---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Amrita Sood"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.
```{r}
getwd()
```

```{r setup}
#1
library(tidyverse)
library(agricolae) 
library(zoo)
library (lubridate)

NTLER_RawData <- 
  read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") 

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes. Intercept and slope is zero indicating no relationship.
Ha: Mean lake temperature recorded during July changes with depths across all lakes
Ho mu = 0
Ha: mu != 0

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTLER_RawData$sampledate<-as.Date(NTLER_RawData$sampledate, format = "%m/%d/%Y")
df <- NTLER_RawData %>% mutate(date = ymd(sampledate)) %>% 
  mutate_at(vars(date), funs(year, month, day))

July_Data <- df %>%
  filter(month=="7")%>%
  select(lakename, year4, daynum, depth, temperature_C)%>%
  na.omit(temperature_C)

#5
Plot.1 <- ggplot(July_Data, aes(x = temperature_C, y = depth, color=depth)) + geom_point()+labs(x="Temperature",y ="Depth", title= "Temperature versus Depth")+ xlim(0, 25) +
 
  geom_smooth(method=lm, color="black")
print(Plot.1)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Negative relationship between depth and temperature. As depth increases the temperature decreases in the lakes in July. The distribution of the points suggests that it is overall linear but there are more data points clustered in the lower depth data points. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

Regression.1 <- lm(data =July_Data, temperature_C ~ depth)
summary(Regression.1)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:73.87 percent variability in temperature is explained by the changes in depth based on degrees of freedom of 9726.
The slope term in our model is saying that for every unit increase in depth, the temperature decreases by 1.946 C.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
Regression.2 <- lm(data = July_Data, temperature_C ~ depth + year4 + daynum)
step(Regression.2)


#10
Model.1 <- lm(data = July_Data, temperature_C ~ depth + year4 + daynum )
summary(Model.1)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: the final set of explanatory variables that AIC suggests is depth, year4 and daynum.
The model explains 74.12% of the observed variance.
Adding more variables increases R-squared value thus making it a better fit and explainign larger variation. Therefore, it is an improvement. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
anova1 <- aov(data = July_Data, temperature_C ~ lakename)
summary(anova1)

anova2 <- lm(data = July_Data, temperature_C ~ lakename)
summary(anova2)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes p value is significant(p-value: < 2.2e-16) and thus there is a signficant difference in mean temperature among the lakes. The linear model does not explain the variability of the model. We can reject the null hypothesis  



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

Plot.2 <- ggplot(July_Data, aes(y = temperature_C, x = depth, color=lakename)) +geom_point(alpha=0.5)+labs(y="Temperature",x ="Depth", title = "Temperature versus Depth for different lakes")+ ylim(0, 35) + geom_smooth(method=lm, se = FALSE, color="black")
print(Plot.2)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(anova1)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:The East Long Lake-Central Long Lake have a difference in mean of -7.39.Hummingbird Lake-Central Long Lake also has a staistically significant difference in mean of -6.89. The following have statisticaly signficant differences in mean. 
Hummingbird Lake-Central Long Lake
Paul Lake-Central Long Lake
Tuesday Lake-Central Long Lake 
Ward Lake-Central Long Lake
West Long Lake-Central Long Lake
East Long Lake-Crampton Lake 
Hummingbird Lake-Crampton Lake
Paul Lake-Crampton Lake 
Peter Lake-Crampton Lake 
Tuesday Lake-Crampton Lake
West Long Lake-Crampton Lake
Paul Lake-East Long Lake
Peter Lake-Crampton Lake 
Peter Lake-Crampton Lake
Tuesday Lake-Crampton Lake 
West Long Lake-Crampton Lake
Paul Lake-East Long Lake
Ward Lake-East Long Lake            
West Long Lake-East Long Lake       
Paul Lake-Hummingbird Lake          
Peter Lake-Hummingbird Lake
Ward Lake-Hummingbird Lake          
Tuesday Lake-Paul Lake             
West Long Lake-Paul Lake           
Tuesday Lake-Peter Lake            
West Long Lake-Peter Lake          
Ward Lake-Tuesday Lake              
West Long Lake-Ward Lake          
>Only Peter and Paul lake have no difference in means which is statistically significant. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: Welch's two sample T test would be useful to see if the two lakes have distinct mean temperatures.


